---
title: "Inference after a custom clustering algorithm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inference after a custom clustering algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`test.clusters.MC` allows to perform inference after any user-supplied clustering algorithm. This vignette
walks through a complete example using [HDBSCAN](https://doi.org/10.1007/978-3-642-37456-2_14) (Hierarchical
Density-Based Spatial Clustering of Applications with Noise). For a user-friendly introduction to HDBSCAN, 
see the [hdbscan package vignette](https://cran.r-project.org/web/packages/dbscan/vignettes/hdbscan.html) or 
the [python package documentation](https://hdbscan.readthedocs.io/en/latest/).

```{r setup}
library(PCIdep)
library(dbscan)   
library(ggplot2)
library(MASS) 
```
---

## 1. Simulate Data

In this example, we simulate a dataset with three "strands" of points drawn from three multivariate Gaussian distributions.
This type of structure can be encountered in real-world data and can be challenging for some clustering algorithms to capture.

```{r simulate-data}
set.seed(42)

n <- 200   # Number of points per cluster
Sigma <- matrix(c(2.5, -0.99, -0.99, 0.45), nrow = 2) # Covariance matrix

# Three Gaussian strands
g1 <- MASS::mvrnorm(n = n, mu = c(-5, 0), Sigma = Sigma)
g2 <- MASS::mvrnorm(n = n, mu = c(0, 0), Sigma = Sigma)
g3 <- MASS::mvrnorm(n = n, mu = c(2, -3), Sigma = Sigma)

X <- as.matrix(rbind(g1, g2, g3))
colnames(X) <- c("x", "y")
df <- as.data.frame(X)

# Plot raw simulated data
ggplot(df, aes(x = x, y = y)) +
  geom_point(alpha = 0.6, size = 1.8) +
  ggtitle("Simulated data") +
  theme_bw()
```
---

## 2. Construct the Cluster Function

We are interested in clustering the data using HDBSCAN and then testing for differences in cluster means between two of the identified clusters. To do so, we need
to construct a function wrapping the HDBSCAN algorithm that can be passed to [`test.clusters.MC`](https://gonzalez-delgado.github.io/PCIdep/reference/test.clusters.MC.html). The wrapper must take as input the data matrix and return a vector of cluster labels.
In the case where the custom clustering algorithm requires the number of clusters to be specified, the wrapper should also take an argument `NC` and pass it to the clustering function. 
In the case of HDBSCAN, the number of clusters is determined automatically, so we do not need to use `NC` in our wrapper. Any other hyperparameters of the clustering algorithm should
be exposed as arguments of the wrapper function so that they can be tuned easily. In the case of HDBSCAN, a key hyperparameter is `minPts` (the minimum cluster size), which we include in our wrapper.

> [`test.clusters.MC`](https://gonzalez-delgado.github.io/PCIdep/reference/test.clusters.MC.html) requires a `cl_fun` argument that follows this
interface:
>
> **`cl_fun(X, NC, ...)` → integer vector of length `nrow(X)`**
>
> * `X` is the data matrix passed to `test.clusters.MC`.
> * `NC` is the number of clusters (if required by the clustering algorithm). If chosen automatically, the `NC` argument can be omitted.
> * The return value must be an integer (or integer-coercible) vector where
>   each element is the cluster label for the corresponding observation.

```{r cluster-fun}
hdbscan_fun <- function(X, minPts = 10) {
  res    <- dbscan::hdbscan(X, minPts = minPts) # minPts controls the minimum cluster size in HDBSCAN.
  labels <- as.integer(res$cluster)
  return(labels)
}
```
**Note:** If the clustering algorithm requires the number of clusters `NC` explicitly, it must be passed to `cl_fun`. For example, a `kmeans` wrapper for [`test.clusters.MC`](https://gonzalez-delgado.github.io/PCIdep/reference/test.clusters.MC.html) can be written as:

```{r cluster-fun-kmeans}
kmeans_fun <- function(X, NC, nstart = 10) {
  fit <- stats::kmeans(X, centers = NC, nstart = nstart)
  as.integer(fit$cluster)
}
```

Let us verify that the `hdbscan_fun` works on our simulated data:

```{r verify-fun}
labels_hdbscan <- hdbscan_fun(X, minPts = 10)
df$hdbscan_cluster <- factor(labels_hdbscan)

ggplot(df, aes(x = x, y = y, colour = hdbscan_cluster)) +
  geom_point(alpha = 0.7, size = 1.8) +
  scale_colour_brewer(palette = "Set1", name = "Cluster\n(0 = noise)") +
  ggtitle("HDBSCAN cluster labels") +
  theme_bw()
```
---

## 3. Run [`test.clusters.MC`](https://gonzalez-delgado.github.io/PCIdep/reference/test.clusters.MC.html)

[`test.clusters.MC`](https://gonzalez-delgado.github.io/PCIdep/reference/test.clusters.MC.html) computes a Monte Carlo based $p$-value to assess whether the means of a pair of clusters
found by the algorithm are significantly different. The function takes the same list of parameters as the [HAC](https://gonzalez-delgado.github.io/PCIdep/reference/test.clusters.hc.html) and [$k$-means](https://gonzalez-delgado.github.io/PCIdep/reference/test.clusters.km.html) functions, with the 
exception of the `cl_fun` and `ndraws` arguments:

Key arguments:

| Argument       | Description |
|----------------|-------------|
| `cluster_fun`  | The wrapper function built in the previous section. |
| `ndraws`       | Number of Monte Carlo iterations (larger → more precise $p$-values). |

> Regarding **covariance estimation**, [the same considerations as in the HAC and $k$-means cases](https://gonzalez-delgado.github.io/PCIdep/articles/hac-km-clustering.html#covariance-estimation) apply. Here, we use sample splitting
to generate an i.i.d. copy of the data for covariance estimation. 

**Use of parallel computation**: The $p$-value is computed via Monte Carlo simulation: the clustering algorithm is run `ndraws` times on perturbed data, and the results are aggregated to approximate the null distribution of the test statistic.
Since each Monte Carlo draw is independent, the computation can be parallelized straightforwardly using the [`future`](https://cran.r-project.org/package=future) package. Simply call `plan()` before running the test and choose the number of workers according to your machine's available cores.

```{r run-test}
set.seed(123)

library(future) 
plan(multisession, workers = 5) # Set parallelization plan (adjust workers according to your machine)

# Compare clusters 1 and 2
res_12 <- test.clusters.MC(X, clusters = c(1, 2), cl_fun = hdbscan_fun, NC = NULL, cl = NULL, ndraws = 500, sample_split = TRUE)
res_12$pvalue

# Compare clusters 2 and 3
res_23 <- test.clusters.MC(X, clusters = c(2, 3), cl_fun = hdbscan_fun, NC = NULL, cl = NULL, ndraws = 500, sample_split = TRUE)
res_23$pvalue

# Compare clusters 1 and 3
res_13 <- test.clusters.MC(X, clusters = c(1, 3), cl_fun = hdbscan_fun, NC = NULL, cl = NULL, ndraws = 500, sample_split = TRUE)
res_13$pvalue
```

The reported $p$-values indicate that the clusters are significantly different.

## Conclusion

This vignette demonstrated how to use `test.clusters.MC` for inference after custom clustering algorithms. 
The key steps are:

1. **Define a wrapper function** that implements your clustering algorithm and returns integer cluster labels.
2. **Run `test.clusters.MC`** with your wrapper function to obtain Monte Carlo-based $p$-values for pairwise cluster comparisons.
3. **Parallelize computation** using the `future` package for faster inference on large datasets.

